```{r setup}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

we will be practicing with `dplyr` functions

```{r}
starwars
head(starwars)
?starwars
```

## Subset rows with `filter`

`filter()` allows us to subset the row of a dataset based on the values

```{r}
##Filter for rows with human starwars char.
filter(starwars, species == "Human")

##save to a new R obj
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

##filter mult. columns at once
filter(starwars, species == "Human", sex == "male")
```

We can use diff. comparison operators to filter:

-   `!` is not

-   `==` is equal

-   `&` is and

-   `|` is or

-   `<` , `>` is less than and grater than

-   `is.na()` returns TRUE if value is NA

```{r}
##return char. with a missing mass
filter(starwars, is.na(mass))
```

```{r}
##how to get all char. that arent human
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values

This is a shortcut for variable_name == vector_of_values[1] \| vector_of_values[2] ...

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") #returns grey because grey isnt in vector of colors

##how would we filter for char with blond or brown hair
filter(starwars, hair_color %in% c("blond", "brown"))
```

grepl("characters_of_interest", variable_name) returns TRUE if the value of variable_name contains the characters of interest

```{r}
filter(starwars, grepl("blond", hair_color))

#How would we get all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can use `<` ,`>` , `<=` , `>=` to compare numeric variable or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on the built-in order\

```{r}
##starwars char is name starting with letters before "D"
filter(starwars, name < "D") # ranking is based on alph. order

str(diamonds)
filter(diamonds, cut > "Good") #rakign is based on levels of the ordered factor "cut"

"D" < "d" #false
"d" < "D" #true
```

between(x, left, right) is a shorthand for x \>= left & x \<= right

```{r}
between(5, 1, 7) #true because 5 is less than 7 and greater than 1
filter(starwars, between(mass, 50, 100))
```

### Practice!

Find all starwars char. that

1.  are missing a hair color
2.  are shorter that 100cm and aren't driods
3.  were born between years 100 and 200
4.  another of your choice

```{r}
filter(starwars, is.na(hair_color)) #1

filter(starwars, height < 100 & species != "Droid") #2

filter(starwars, between(birth_year, 100, 200)) #3

filter(starwars, hair_color == "black") #4
```

Order rows with `arrange`

`arrange` reorders rows based o values of the columns

```{r}
#order by char names
arrange(starwars, name)
#order of char val: aAbBcCd

##order by species and birthyear
arrange(starwars, species, birth_year)

##use desc() to get the birthyears in descending order
arrange(starwars, species, desc(birth_year))
```

Practice!

1.  arrange the starwars dataset from tallest to shortest
2.  arrange the starwars dataset from shortest to tallest by species
3.  Alphabetize the Star Wars characters by their hair color then eye color

```{r}
arrange(starwars, desc(height)) #1
arrange(starwars, height, species) #2 order by height, tiebreaker is alph species
arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors 
```

We can also select multiple columns in a row

```{r}
#all colums between name and eyecolor inclusive
select(starwars, name:eye_color)
```

We can also remove columns by name using a -

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything())#puts species first than everything else

##alphabetize columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain with "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

Practice

1.  Select each character homeworld and species. What question might this subset of var. help answer
2.  brainstorm as many ways a possible to select haircolor, eye color, skin color and birthyear

```{r}
select(starwars, c(name, homeworld, species))
#question can answer the variation of species between homeworlds

select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, c(hair_color, skin_color, eye_color, birth_year))
select(starwars, hair_color:birth_year)
select(starwars, c(hair_color:birth_year))
select(starwars, contains("color"), birth_year)
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create multiple variable at once and refer to variables that we just created

```{r}
mutate(starwars_small,
       height_m = height/100, 
       bmi = mass/height_m^2)

##only keep new var
transmute(starwars_small,
       height_m = height/100, 
       bmi = mass/height_m^2)
```

summary stats. give one value for the entire column: `sum()`, `mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group by variable using `.by`

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.rm = T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small, 
       mass = as.character(mass))
```

Practice:

1.  convert mass to pounds

```{r}
mutate(starwars_small, 
       species_pounds = mass*2.205)
```

## Summarize data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#Group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the column you group by and the columns you create.

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = TRUE),
          count = n(),
          .by = species)

#group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = TRUE),
          count = n(),
          .by = c(species,sex))
```

## Renaming

rename(data, new_name = old_name)

```{r}
starwars
new_starwars <- (starwars,
                 birthYear = birth_year)
new_starwars
```
